---
title: "Estimating population usage from input distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating population usage from input distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(netz)
library(ggplot2)
```

In [malariasimulation](https://mrc-ide.github.io/malariasimulation/) we can implement the [bed net intervention](https://mrc-ide.github.io/malariasimulation/articles/VectorControl.html) by specifying
the % of the population who are distributed a bed net (and assumed to use it) at
given time points.

Often it is helpful to understand the resulting population levels of bed net usage from a set of 
bed net distributions over time. The 5 of individuals protected by a net at any time point is a 
function of the cumulative distributions that have occurred, the durations of net retention and the
correlation of recipients between rounds.

In netz, we can simulate population usage and examine how varying distribution size and frequency,
net retention and recipient correlation impacts our estimates.

Let's start with a simple distribution model:

* Distribution cycle: A mass distribution once every 3 years
* Distribution size: 70% of the population covered
* Net retention: Assume net retention decays exponentially with a half life of 5 years
* Correlation: Recipients are correlated each round

```{r, fig.width = 7, fig.height=4}
year <- 365
timesteps <- 30 * year
population <- 5000

distribution <- rep(0.3, 10)
distribution_timesteps <- year * seq(1, 30, 3)

net_hl <- 5 * year
net_loss <- exp(-(1 / net_hl) * 1:timesteps)

correlation <- 1

pu <- population_usage(
    net_loss = net_loss,
    population = population,
    timesteps = timesteps,
    distribution = distribution,
    distribution_timesteps = distribution_timesteps,
    rho = correlation)

pd <- data.frame(t = 1:timesteps, usage = pu)

ggplot(pd, aes(x = t, y = usage)) +
  geom_line() +
  ylim(0, 0.5) +
  theme_bw()

```

We can see that population usage initially rises with the first distribution population before reaching an equilibrium as following distibutions replace the nets of those who already own one.

In the simple case of regular distributions and exponential net loss and correlation we have the
analytical solution for the average population level usage.

```{r, fig.width = 7, fig.height=4}
analytical <- exp_loss_equilibrium(distribution = 0.3, distribution_frequency = 3, net_loss_hl = 5)

ggplot(pd, aes(x = t, y = usage)) +
  geom_line() +
  geom_hline(yintercept = analytical, lty = 2, col = "darkred") +
  ylim(0, 0.5) +
  theme_bw()
```

More frequent distribution will lead to net usage to build to higher levels

```{r, fig.width = 7, fig.height=4}
distribution2 <- rep(0.3, 30)
distribution_timesteps2 <- year * seq(1, 30, 1)

pu <- population_usage(
    net_loss = net_loss,
    population = population,
    timesteps = timesteps,
    distribution = distribution2,
    distribution_timesteps = distribution_timesteps2,
    rho = correlation)

pd <- data.frame(t = 1:timesteps, usage = pu)

analytical <- exp_loss_equilibrium(distribution = 0.3, distribution_frequency = 1, net_loss_hl = 5)

ggplot(pd, aes(x = t, y = usage)) +
  geom_line() +
  geom_hline(yintercept = analytical, lty = 2, col = "darkred") +
  ylim(0, 0.5) +
  theme_bw()

```
Random recipients will lead to higher net usage as well as each additional distribution has the chance to reach an individual who has not received a net previously

```{r, fig.width = 7, fig.height=4}
correlation2 <- 0

pu <- population_usage(
    net_loss = net_loss,
    population = population,
    timesteps = timesteps,
    distribution = distribution,
    distribution_timesteps = distribution_timesteps,
    rho = correlation2)

pd <- data.frame(t = 1:timesteps, usage = pu)

ggplot(pd, aes(x = t, y = usage)) +
  geom_line() +
  ylim(0, 0.5) +
  theme_bw()
```

We may also need to estimate population usage when distributions are not regular

```{r, fig.width = 7, fig.height=4}
distribution3 <- c(0.1, 0.1, 0.35, 0.1, 0.23, 0.2, 0.32, 0.1, 0.01, 0.2)
distribution_timesteps3 <- seq(1, 30, 3) * year
  
pu <- population_usage(
    net_loss = net_loss,
    population = population,
    timesteps = timesteps,
    distribution = distribution3,
    distribution_timesteps = distribution_timesteps3,
    rho = correlation)

pd <- data.frame(t = 1:timesteps, usage = pu)

ggplot(pd, aes(x = t, y = usage)) +
  geom_line() +
  ylim(0, 0.5) +
  theme_bw()
```

Finally, it may be useful to explore other net-loss functions, for example a sigmoidal decay:

```{r, fig.width = 7, fig.height=4}
net_hl2 <- 5 * year
k <- - 1/150
t0 <- 365
net_loss2 <- 1 / (1 + exp(-k * (1:timesteps - t0)))

pu <- population_usage(
    net_loss = net_loss2,
    population = population,
    timesteps = timesteps,
    distribution = distribution,
    distribution_timesteps = distribution_timesteps,
    rho = correlation)

pd <- data.frame(t = 1:timesteps, usage = pu)

ggplot(pd, aes(x = t, y = usage)) +
  geom_line() +
  ylim(0, 0.5) +
  theme_bw()

```